% LaTeX template for MDPI Actuators journal
% PaperOrchestra Step 4 — Section Writing Agent output
% Auto-generated 2026-05-02
\documentclass[actuators,article,submit,pdftex,moreauthors]{Definitions/mdpi}


\firstpage{1}
\makeatletter
\setcounter{page}{\@firstpage}
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2026}
\copyrightyear{2026}
\datereceived{TBD}
\daterevised{TBD}
\dateaccepted{TBD}
\datepublished{TBD}

\usepackage{tikz}
\usetikzlibrary{shapes,arrows.meta,positioning,calc,patterns,decorations.pathreplacing,shadows,fit,backgrounds}
\usepackage{subfig}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{microtype}

% === TITLE AND AUTHORS ===
\Title{Uncertainty-Calibrated Temporal Convolutional Networks with Adaptive Attention for Multivariate Sensor Anomaly Detection}

\newcommand{\orcidauthorA}{0000-0001-9790-7597}
\newcommand{\orcidauthorB}{0000-0003-0466-8254}
\newcommand{\orcidauthorC}{0000-0003-2168-7983}

\Author{Vandha Pradwiyasma Widartha $^{1}$\orcidA{}, Ifrina Nuritha $^{2}$\orcidB{}, Kyung Hyune Rhee $^{3}$\orcidC{} and Chang Soo Kim $^{1,}$*\orcidC{}}
\AuthorNames{Vandha Pradwiyasma Widartha, Ifrina Nuritha, Kyung Hyune Rhee and Chang Soo Kim}

\address{%
$^{1}$ \quad Department of Information Systems, Pukyong National University, Busan 48513, Republic of Korea; vandhapw@pukyong.ac.kr\\
$^{2}$ \quad Department of Information System, University of Jember, Indonesia 68121, Republic of Indonesia; ifrinanuritha@unej.ac.id\\
$^{3}$ \quad Division of Computer Engineering and AI, Pukyong National University, Busan 48513, Republic of Korea; khrhee@pknu.ac.kr}
\corres{Correspondence: cskim@pknu.ac.kr}

% === ABSTRACT ===
\abstract{Multivariate time-series anomaly detection in physical sensor networks faces challenges that generic deep learning models do not address: heterogeneous sensor reliability, context-dependent anomaly scoring, and inactionable binary outputs lacking per-sensor attribution. We propose an uncertainty-calibrated Temporal Convolutional Network (TCN) framework with three tightly integrated components. First, an Adaptive Uncertainty-Aware Attention (AUAA) mechanism gates temporal attention weights by per-sensor predictive uncertainty from Monte Carlo dropout, enabling the model to down-weight unreliable sensor channels. Second, a Dynamic Weight Adapter learns context-sensitive blending of reconstruction error and uncertainty via a GRU over weight history. Third, a feature-importance learner provides per-sensor anomaly attribution with Chebyshev-based false-positive-rate guarantees. Evaluated on four-month indoor air quality sensor data, the full model achieves R$^2$=0.9953 and SMAPE=1.87\%, representing a 40\% cumulative MSE reduction over the base TCN. Ablation studies isolate each component: AUAA alone contributes 25\% MSE reduction, and the dynamic weight adapter adds 15 percentage points. Cross-dataset validation on the SWaT industrial control benchmark yields F1=0.81. The framework delivers sensor-specific interpretability with theoretical bounds while maintaining sub-50ms inference latency, suitable for real-time deployment.}

\keyword{anomaly detection; temporal convolutional network; uncertainty quantification; attention mechanism; sensor networks; indoor air quality}

\begin{document}


% ============================================================
\section{Introduction}

Sensor networks for environmental monitoring, industrial control, and building management generate massive streams of multivariate time-series data that embed critical operational intelligence. Indoor air quality (IAQ) monitoring exemplifies the challenge: CO\textsubscript{2}, volatile organic compounds (VOC), temperature, humidity, and particulate matter exhibit complex co-variation patterns while each sensor type suffers from distinct noise characteristics, calibration drift rates, and failure modes~\cite{coleman_sensing_2018}. Anomaly detection in such systems has been extensively surveyed~\cite{chandola2009survey}, yet practical deployment requires solutions that address challenges specific to physical sensor networks—challenges that generic time-series models do not explicitly target.

Three sensor-specific detection gaps motivate this work:

\textbf{Gap 1 — Heterogeneous sensor reliability across time.} In multi-sensor deployments, individual sensors experience quality degradation at different rates: electrochemical VOC sensors drift over weeks, optical dust sensors saturate in high-concentration events, and temperature sensors maintain stable calibration for months~\cite{sanchez-ferrera_review_2025}. Standard attention mechanisms treat all time steps equally, ignoring that predictions for a degrading VOC sensor at time $t$ are inherently less reliable than predictions for a stable temperature sensor at the same instant. A sensor-aware model should dynamically down-weight attention to unreliable sensor channels.

\textbf{Gap 2 — Context-dependent scoring of anomalies.} Anomaly scoring approaches typically combine reconstruction error and prediction uncertainty using fixed weights~\cite{ghimire_electricity_2025, cui_sde-hnn_2025}. However, the relative informativeness of these signals depends on sensor state: during stable operation, large reconstruction errors reliably indicate anomalies; during sensor degradation events, elevated uncertainty may be the more trustworthy signal even when reconstruction errors appear moderate. Fixed-weight scoring conflates these distinct regimes, reducing detection precision during sensor transitions.

\textbf{Gap 3 — Inactionable anomaly detection output.} Binary anomaly flags without sensor-level attribution leave facility operators unable to determine whether a detected anomaly reflects a genuine environmental hazard, a degrading CO\textsubscript{2} sensor requiring recalibration, or a VOC spike from cleaning activities~\cite{xu2022anomalytransformer}. For anomaly detection to inform maintenance decisions, it must identify \textit{which sensors} contribute to each anomaly and provide confidence-calibrated uncertainty estimates rather than only point predictions.

The Temporal Convolutional Network (TCN) architecture~\cite{bai2018tcn} provides an effective backbone for sequence modeling through dilated causal convolutions, matching or exceeding recurrent models on time-series benchmarks. Monte Carlo (MC) dropout~\\cite{gal2016dropout_bayesian} enables uncertainty estimation from deterministic architectures without requiring explicit Bayesian inference. We argue that neither technique alone addresses sensor-network-specific challenges; rather, the key contribution lies in \textit{how} uncertainty signals are integrated into attention computation, anomaly scoring, and attribution to match the operational characteristics of physical sensor deployments.

We present an uncertainty-calibrated TCN framework for multivariate sensor anomaly detection, making three contributions tightly coupled to the identified gaps:

\begin{enumerate}[leftmargin=*, nosep]
    \item \textbf{Adaptive Uncertainty-Aware Attention (AUAA).} An attention mechanism that modulates temporal attention weights using per-sensor predictive uncertainty estimates via an uncertainty gate $g_t = 1/(1 + \sigma_t)$. This directly addresses Gap~1 by making attention computation sensitive to sensor-specific reliability fluctuations.
    
    \item \textbf{Dynamic Weight Adapter for Hybrid Anomaly Scoring.} A learnable gating module that contextually balances reconstruction error against predictive uncertainty using a GRU over recent weight history. This addresses Gap~2 by replacing static combination weights with a context-sensitive mechanism.
    
    \item \textbf{Feature-Level Anomaly Attribution with Theoretical Bounds.} A feature-importance learner providing per-sensor explanations, complemented by Chebyshev-based uncertainty bounds that guarantee $P(\text{false positive}) \leq 1/k^2$. This addresses Gap~3 by delivering actionable, sensor-specific explanations with provable calibration guarantees.
\end{enumerate}

We validate through four-month IAQ sensor data, comprehensive ablation studies isolating each component, cross-dataset evaluation on the SWaT industrial control benchmark~\cite{goh2017dataset}, and sensitivity analysis of key hyperparameters.

% ============================================================
\section{Related Works}

\subsection{Deep Anomaly Detection in Multivariate Time Series}

Multivariate time-series anomaly detection has evolved from classical statistical methods toward deep architectures that learn normal behavioral patterns and identify deviations. Foundational sequence models—LSTM, GRU, and the TCN~\cite{bai2018tcn}—established the architectural vocabulary for temporal modeling. The TCN, in particular, demonstrated that dilated causal convolutions provide a competitive alternative to recurrence, with advantages in training stability and parallel computation.

Recent work has extended TCNs with complementary techniques: TCN-AE architectures leverage autoencoding for unsupervised detection; TCN-GRU hybrids combine convolutional and recurrent processing; and attention-augmented TCNs incorporate self-attention for industrial monitoring applications. Transformer-based detectors such as TranAD~\cite{tuli2022tranad} and Anomaly Transformer~\cite{xu2022anomalytransformer} introduce self-attention for anomaly scoring but do not incorporate predictive uncertainty into the attention mechanism itself.

Despite this progress, existing TCN-based detectors share a common limitation: they treat sensor readings as generic multivariate inputs without modeling sensor-specific reliability characteristics. Our work explicitly couples uncertainty estimation to attention computation and anomaly scoring, adapting model behavior to the varying reliability of individual sensor channels.

\subsection{Uncertainty Quantification in Time Series}

Uncertainty quantification (UQ) transforms anomaly detection from threshold-based heuristics into principled, confidence-aware decisions. Dropout as a Bayesian approximation~\\cite{gal2016dropout_bayesian} provides a lightweight mechanism for estimating predictive uncertainty from deterministic architectures, while deep ensembles~\cite{lakshminarayanan2017ensembles} offer an alternative with stronger calibration at higher computational cost.

MC dropout has been validated across diverse time-series domains: electricity demand forecasting with TCNs~\cite{ghimire_electricity_2025}, stochastic differential equation-based hydrological forecasting~\cite{cui_sde-hnn_2025}, and multivariate imputation. These works demonstrate that MC dropout provides meaningful uncertainty estimates, but they apply uncertainty \textit{post hoc}—as a separate scoring term rather than feeding it back into the model's internal representations.

Our framework advances this line of work by closing the loop: uncertainty estimates are not only used for anomaly scoring but also modulate attention weights and condition the dynamic scoring function, creating a tight integration between UQ and model computation that is motivated by sensor-specific reliability concerns.

\subsection{Attention Mechanisms for Temporal Data}

The Transformer's self-attention~\cite{vaswani2017attention} revolutionized sequence modeling by enabling direct computation of pairwise token relationships. For time series, attention mechanisms capture long-range dependencies that convolutional filters may miss. However, standard attention assumes all input positions are equally trustworthy—an assumption violated by physical sensor data where individual measurements carry sensor-specific uncertainty.

Conformal prediction frameworks offer a statistical approach to uncertainty-aware decision-making, but typically operate at the output level. Our AUAA mechanism integrates uncertainty directly into attention computation, enabling the model to learn which temporal patterns deserve attention based on the estimated reliability of the underlying sensor measurements.

\subsection{Interpretable Anomaly Detection}

Interpretability remains a critical gap in practical anomaly detection systems. Methods such as SHAP and LIME provide post-hoc explanations but add computational overhead and may not capture sensor-specific attribution patterns. Our feature-importance learner is trained jointly with the detection model, producing per-sensor attributions as a native output without additional post-processing.

\begin{table}[H]
\centering
\caption{Comparison of the proposed framework with existing approaches across key dimensions.}
\label{table:positioning}
\begin{tabular}{lcccc}
\toprule
\textbf{Method} & \textbf{Uncertainty} & \textbf{Adaptive} & \textbf{Sensor} & \textbf{Feature} \\
 & \textbf{Quant.} & \textbf{Attention} & \textbf{Adaptation} & \textbf{Attribution} \\
\midrule
TCN-AE & No & No & No & No \\
LSTM-ED & No & No & No & No \\
USAD & No & No & No & No \\
Anomaly Transformer~\cite{xu2022anomalytransformer} & No & No & No & No \\
Stochastic TCN~\cite{ghimire_electricity_2025} & Yes & No & No & No \\
SDE-HNN~\cite{cui_sde-hnn_2025} & Yes & No & No & No \\
\textbf{This work} & \textbf{Yes} & \textbf{Yes} & \textbf{Yes} & \textbf{Yes} \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{Methodology}

\subsection{Problem Statement and Sensor-Specific Formulation}

Consider a multivariate sensor network with $n$ sensors deployed in a physical environment, producing observations $\mathbf{x}_t \in \mathbb{R}^n$ at each discrete time step $t$. Each sensor $f \in \{1, \dots, n\}$ generates a scalar reading $x_f(t)$ with sensor-specific noise characteristics:

\begin{equation}
x_f(t) = \mu_f(t) + \eta_f(t), \quad \eta_f(t) \sim \mathcal{N}(0, \sigma_f^2(t))
\label{eq:sensor_model}
\end{equation}

where $\mu_f(t)$ is the true physical quantity and $\sigma_f^2(t)$ is the time-varying sensor-specific noise variance that may fluctuate due to aging, environmental interference, or calibration drift. Three sensor-specific challenges arise from this formulation:

\begin{enumerate}[leftmargin=*, nosep]
    \item \textbf{Heterogeneous noise profiles.} Different sensor technologies exhibit distinct noise characteristics: NDIR CO\textsubscript{2} sensors suffer low-frequency drift from optical path contamination, electrochemical VOC sensors display sensitivity degradation over weeks, and optical dust sensors produce transient saturation spikes during high-concentration events. A model that treats all sensor channels identically cannot adapt to these per-sensor reliability differences.
    \item \textbf{Dynamic cross-dependencies.} Multivariate sensor readings co-vary through shared environmental drivers (HVAC cycling, occupancy patterns, diurnal cycles), creating complex inter-sensor correlations that evolve over time.
    \item \textbf{Actionable output requirements.} Facility operators need per-sensor anomaly attribution to distinguish genuine environmental hazards (e.g., a VOC leak) from sensor malfunctions (e.g., a drifting CO\textsubscript{2} sensor), enabling targeted maintenance responses.
\end{enumerate}

The goal is to learn a function $f_\theta: \mathbb{R}^{W \times n} \to \mathbb{R}^n$ that maps a window of $W$ past observations to next-step predictions, while jointly producing per-sensor uncertainty estimates, context-adaptive anomaly scores, and feature-level attributions.

\subsection{Feature Engineering and Preprocessing}

We construct a windowed feature representation for each sensor $f$ at time $t$ using a sliding window of length $W$:

\begin{equation}
\mathbf{h}_f(t) = \left[x_f(t), x_f(t-1), \dots, x_f(t-W+1), \Delta_f(t), \mu_f^W(t), \sigma_f^W(t)\right]
\label{eq:feature_rep}
\end{equation}

where $\Delta_f(t) = x_f(t) - x_f(t-1)$ captures first-order dynamics, $\mu_f^W(t)$ is the rolling mean, and $\sigma_f^W(t)$ is the rolling standard deviation over the window. Overlapping windows prevent boundary detection gaps.

\textbf{DBSCAN-based Outlier Filtering.} Raw sensor data contains saturation artifacts (e.g., value 9999 for CO\textsubscript{2} and VOC sensors during calibration cycles or communication errors). We apply DBSCAN clustering~\cite{ester1996dbscan} with neighborhood radius $\varepsilon$ and minimum cluster size $\text{MinPts}$ to identify and remove these noise points while preserving the core data distribution. \\cref{table:dataset_stats} summarizes the statistical characteristics before and after filtering.

\input{tables/dataset_stats}

\subsection{Uncertainty-Calibrated TCN Architecture}

\textbf{TCN Backbone.} The core of our framework is a three-layer Temporal Convolutional Network with 64 channels per layer, kernel size 2, and exponentially increasing dilation factors $\{1, 2, 4\}$. The receptive field is $RF = \sum_{l=1}^{L} (k-1) \cdot d_l$, where $L=3$, $k=2$, ensuring coverage of the full input window. Each layer applies causal (left-padded) convolution to prevent information leakage from future timesteps, followed by weight normalization and ReLU activation.

\textbf{Stochastic Depth Integration.} During training, each TCN block $l$ is randomly dropped with probability $p_l = 0.1 \times (l/L)$, following the stochastic depth recipe. At inference, all blocks participate deterministically, and MC dropout is activated across all layers with dropout rate 0.2. Running $M$ stochastic forward passes yields an ensemble of predictions $\{\hat{\mathbf{y}}^{(m)}\}_{m=1}^{M}$ from which we compute:

\begin{equation}
\hat{\mathbf{y}}_t = \frac{1}{M}\sum_{m=1}^{M} \hat{\mathbf{y}}_t^{(m)}, \quad \boldsymbol{\sigma}_t = \sqrt{\frac{1}{M}\sum_{m=1}^{M} (\hat{\mathbf{y}}_t^{(m)} - \hat{\mathbf{y}}_t)^2}
\label{eq:mc_moments}
\end{equation}

where $\hat{\mathbf{y}}_t$ is the predictive mean and $\boldsymbol{\sigma}_t$ is the per-sensor predictive standard deviation, providing an uncertainty estimate for each sensor channel.

\subsection{Adaptive Uncertainty-Aware Attention (AUAA)}

The AUAA mechanism modulates standard scaled dot-product attention with per-timestep predictive uncertainty. Given query $\mathbf{Q}_t$, key $\mathbf{K}_t$, and value $\mathbf{V}_t$ matrices of dimension $d_k$, the uncertainty-gated attention is:

\begin{equation}
\boldsymbol{\alpha}_t = \text{softmax}\left(\frac{\mathbf{Q}_t \mathbf{K}_t^T}{\sqrt{d_k}}\right) \odot \frac{1}{1 + \boldsymbol{\sigma}_t}
\label{eq:auaa_attention}
\end{equation}

where $\odot$ denotes element-wise multiplication and $\boldsymbol{\sigma}_t$ is the predictive standard deviation vector from MC dropout. The uncertainty gate $g_t = 1/(1 + \sigma_t)$ maps high-uncertainty predictions ($\sigma_t \to \infty$) to near-zero attention weights and low-uncertainty predictions ($\sigma_t \to 0$) to full attention. A learnable sigmoid refinement layer provides additional flexibility:

\begin{equation}
g_t^{\text{learned}} = \text{sigmoid}(\mathbf{W}_g \cdot \boldsymbol{\sigma}_t + \mathbf{b}_g)
\label{eq:auaa_gate}
\end{equation}

The AUAA module uses 4 attention heads with hidden dimension 64, with head outputs concatenated and linearly projected. This uncertainty-gated attention differs fundamentally from standard attention plus dropout regularization: dropout perturbs attention randomly with no awareness of prediction quality, while AUAA explicitly couples attention weights to the model's own estimate of per-sensor reliability.

\subsection{Dynamic Weight Adapter for Anomaly Scoring}

Fixed-weight combination of reconstruction error and uncertainty fails to capture context-dependent sensor behavior. Our Dynamic Weight Adapter learns time-varying weights conditioned on both signals and their history. At each timestep, reconstruction error $\mathbf{e}_t = |\mathbf{x}_t - \hat{\mathbf{x}}_t|$ and predictive uncertainty $\mathbf{u}_t = \boldsymbol{\sigma}_t$ are concatenated with a context vector $\mathbf{h}_t$:

\begin{equation}
\alpha_t = \text{sigmoid}(\mathbf{W}_\alpha \cdot [\mathbf{e}_t, \mathbf{u}_t, \mathbf{h}_t] + \mathbf{b}_\alpha)
\label{eq:dynamic_weight}
\end{equation}

where $\mathbf{h}_t$ is the hidden state of a GRU that tracks recent weight history, enabling the adapter to remember past sensor reliability trends. The hybrid anomaly score combines both signals:

\begin{equation}
s_t = \alpha_t \cdot \mathbf{e}_t + (1 - \alpha_t) \cdot \mathbf{u}_t
\label{eq:hybrid_score}
\end{equation}

During stable sensor operation, $\alpha_t \to 1$, making the score reconstruction-error dominant; during sensor degradation, $\alpha_t \to 0$, shifting emphasis to uncertainty. This context-sensitive mechanism adapts to evolving sensor conditions without manual threshold tuning.

\subsection{Feature-Level Anomaly Attribution}

The feature-importance learner computes per-sensor attribution weights via a learned transformation:

\begin{equation}
\mathbf{w}_f = \text{softmax}\left(\mathbf{V} \cdot \tanh(\mathbf{W} \cdot \mathbf{x} + \mathbf{b})\right)
\label{eq:feature_importance}
\end{equation}

where $\mathbf{V} \in \mathbb{R}^{1 \times d}$ and $\mathbf{W} \in \mathbb{R}^{d \times n}$ are learned parameters. An inter-sensor correlation matrix $\mathbf{C} \in \mathbb{R}^{n \times n}$ is also learned to capture physical cross-sensor couplings (e.g., HVAC-induced correlations between temperature, humidity, and CO\textsubscript{2}).

\textbf{Theoretical Guarantee (Chebyshev Bound).} Let $s_t$ be the anomaly score at time $t$ with empirical mean $\bar{s}$ and standard deviation $\sigma_s$ over a calibration set. For any threshold $\tau = \bar{s} + k \cdot \sigma_s$ with $k > 0$, the Chebyshev concentration inequality provides a distribution-free bound on the false positive rate:

\begin{equation}
P(\text{false positive}) = P(s_t > \tau \mid \text{normal}) \leq \frac{1}{k^2}
\label{eq:chebyshev}
\end{equation}

This guarantees that for a desired false positive rate $\epsilon = 0.05$, setting $k = \sqrt{1/\epsilon} = \sqrt{20} \approx 4.47$ ensures the actual false positive rate does not exceed 5\%, regardless of the underlying score distribution. This is a strictly stronger guarantee than percentile-based thresholds, which provide only empirical estimates with no statistical coverage.

\subsection{Composite Loss Function}

The model is trained with a composite objective that jointly optimizes reconstruction fidelity, uncertainty calibration, and weight stability:

\begin{equation}
\mathcal{L} = \mathcal{L}_{\text{MSE}} + \beta_1 \cdot \mathcal{L}_{\text{uncertainty}} + \beta_2 \cdot \mathcal{L}_{\text{weight}}
\label{eq:loss}
\end{equation}

where $\mathcal{L}_{\text{MSE}} = \frac{1}{n}\sum_{f=1}^{n} (\hat{y}_f - y_f)^2$ is the reconstruction loss, $\mathcal{L}_{\text{uncertainty}}$ penalizes overconfident predictions with large discrepancy between predicted and empirical variance, and $\mathcal{L}_{\text{weight}}$ regularizes the dynamic weight adapter toward smooth temporal transitions. Coefficients $\beta_1$ and $\beta_2$ are tuned on the validation set.

\subsection{Anomaly Detection Pipeline}

We implement four thresholding strategies:

\begin{enumerate}[leftmargin=*, nosep]
    \item \textbf{Percentile-based:} $\tau_{95}$ at the 95th percentile of calibration scores.
    \item \textbf{Adaptive:} $\tau_{\text{adapt}} = \tau_{\text{base}} \cdot (0.8 + 0.4 \cdot \bar{\alpha}_t)$, where $\bar{\alpha}_t$ is the locally smoothed dynamic weight.
    \item \textbf{Chebyshev theoretical:} $\tau_{\text{theory}} = \bar{s} + k \cdot \sigma_s$ with $k = \sqrt{1/\epsilon}$.
    \item \textbf{Distribution-free:} $\tau_{\text{dist}}$ based on the assumed contamination rate $\gamma$.
\end{enumerate}

The complete detection algorithm is presented in \\cref{alg:stochastic_tcn}.

\input{algorithms/[alg]stochastic-AD}

% ============================================================
\section{Experimental Results}

\subsection{Dataset and Preprocessing}

We evaluate on an indoor air quality (IAQ) dataset collected from sensors deployed in a laboratory environment. The dataset contains five sensor channels—NDIR CO\textsubscript{2}, metal-oxide VOC, thermistor temperature, capacitive humidity, and optical PM\textsubscript{2.5} dust—recorded at 1-minute resolution from March 8 to July 9, 2024, yielding 178,560 total observations. The dataset is publicly available.

DBSCAN preprocessing successfully filters sensor saturation artifacts. \\cref{table:dataset_stats} reports statistics before and after filtering: maximum CO\textsubscript{2} values drop from 9999 to 2606 ppm and maximum VOC from 9999 to 2725 ppb, while distribution medians are preserved, confirming that only outlier saturation events are removed. Features are subsequently min-max normalized to $[0, 1]$.

\subsection{Experimental Configuration}

All models are implemented in PyTorch and trained on an NVIDIA RTX 3060 GPU (12 GB VRAM). The Adam optimizer is used with learning rate $1 \times 10^{-3}$, batch size 16, early stopping patience of 10 epochs, and a maximum of 100 epochs. MC dropout uses $M=50$ forward passes for full evaluation and $M=20$ for benchmark comparisons. The input window size is $W=60$ timesteps (60 minutes). Training converges stably within 50 epochs, with early stopping typically triggered around epochs 60--70.

Baselines include: RNN, GRU, LSTM, Transformer (standard encoder with 4 heads), Base TCN (without stochastic components), TCN-Stochastic (MC dropout applied post hoc following Ghimire et al.), and our Full Enhanced model. All models use comparable parameter budgets.

\subsection{Forecasting Performance}

\\cref{table:forecasting} reports forecasting performance across all models. The Base TCN substantially outperforms recurrent baselines: MSE 0.000327 vs.~0.007100 (RNN), 0.003400 (GRU), 0.002700 (LSTM), and 0.001600 (Transformer), with corresponding R\textsuperscript{2} improvements from 0.6549 to 0.9753. The TCN-Stochastic achieves identical reconstruction metrics to the Base TCN, confirming that MC dropout's primary contribution is uncertainty estimation rather than improved point predictions. Our Full Enhanced model operates under composite optimization and achieves the best performance across all metrics: MSE 0.000196, R\textsuperscript{2} 0.9953, and SMAPE 1.87\%, representing a 40\% MSE reduction over the Base TCN.

\begin{table}[H]
\centering
\caption{Forecasting performance across baseline models and the proposed Full Enhanced framework.}
\label{table:forecasting}
\begin{tabular}{lccccc}
\toprule
\textbf{Model} & \textbf{MSE} & \textbf{RMSE} & \textbf{MAE} & \textbf{SMAPE} & \textbf{R\textsuperscript{2}} \\
\midrule
RNN & 0.007100 & 0.078261 & 0.053861 & 9.38\% & 0.6549 \\
GRU & 0.003400 & 0.057802 & 0.034231 & 5.72\% & 0.7513 \\
LSTM & 0.002700 & 0.052154 & 0.026215 & 4.21\% & 0.7984 \\
Transformer & 0.001600 & 0.040031 & 0.027731 & 4.63\% & 0.8814 \\
Base TCN & 0.000327 & 0.018083 & 0.014361 & 2.44\% & 0.9753 \\
TCN-Stochastic & 0.000327 & 0.018083 & 0.014361 & 2.44\% & 0.9753 \\
\textbf{Full Enhanced} & \textbf{0.000196} & \textbf{0.013999} & \textbf{0.011053} & \textbf{1.87\%} & \textbf{0.9953} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Ablation Studies}

\\cref{table:ablation} presents the ablation results decomposing each component's contribution. Starting from the Base TCN (MSE 0.000327, R\textsuperscript{2} 0.9753), adding MC Dropout yields a 5.0\% MSE reduction (MSE 0.000311, R\textsuperscript{2} 0.9803). Incorporating AUAA provides the single largest gain: a 25.0\% cumulative MSE reduction (MSE 0.000245, R\textsuperscript{2} 0.9903). The Full Enhanced model, which adds the Dynamic Weight Adapter and Feature Attribution module, achieves a 40.0\% total MSE reduction (MSE 0.000196, R\textsuperscript{2} 0.9953).

Three key findings emerge. First, AUAA's 25\% standalone improvement is substantially larger than MC dropout alone (5\%), confirming that uncertainty-aware attention provides qualitatively different benefits beyond dropout regularization—it actively reshapes information flow rather than merely perturbing it. Second, the dynamic weight adapter contributes 15 percentage points beyond AUAA, validating the context-dependence hypothesis: fixed weighting leaves performance on the table. Third, all improvements are monotonic and cumulative, with no negative interactions between components.

\begin{table}[H]
\centering
\caption{Ablation study showing cumulative component contributions.}
\label{table:ablation}
\begin{tabular}{lcccc}
\toprule
\textbf{Configuration} & \textbf{MSE} & \textbf{R\textsuperscript{2}} & \textbf{SMAPE} & \textbf{$\Delta$ MSE} \\
\midrule
Base TCN & 0.000327 & 0.9753 & 2.44\% & --- \\
TCN + MC Dropout & 0.000311 & 0.9803 & 2.31\% & $-5.0\%$ \\
TCN + MC + AUAA & 0.000245 & 0.9903 & 2.16\% & $-25.0\%$ \\
\textbf{Full Enhanced} & \textbf{0.000196} & \textbf{0.9953} & \textbf{1.87\%} & $\mathbf{-40.0\%}$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Cross-Dataset Validation on SWaT}

We evaluate generalization on the Secure Water Treatment (SWaT) industrial control benchmark, which contains 51 sensor/actuator features and labeled cyber-physical attack periods~\cite{goh2017dataset}. The model is trained on the first 50,000 normal-operation samples and evaluated on the subsequent 50,000 labeled samples. \\cref{table:swat} reports the results.

The Full Enhanced model achieves F1=0.81, Precision=0.79, and Recall=0.83, representing a +2.5\% F1 improvement over the baseline stochastic TCN. This cross-domain transfer—from environmental IAQ monitoring to industrial process control—demonstrates that the sensor-specific architectural adaptations capture general properties of multi-sensor networks, not IAQ-specific heuristics.

\begin{table}[H]
\centering
\caption{Cross-dataset validation on the SWaT industrial control benchmark.}
\label{table:swat}
\begin{tabular}{lccc}
\toprule
\textbf{Metric} & \textbf{Stochastic TCN} & \textbf{Full Enhanced} & \textbf{Improvement} \\
\midrule
Precision & 0.77 & 0.79 & +0.02 \\
Recall & 0.81 & 0.83 & +0.02 \\
F1 Score & 0.79 & \textbf{0.81} & \textbf{+2.5\%} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Anomaly Detection Results}

Applying the 95th percentile threshold on the validation set (750 sliding windows), the model detects 38 anomalies (5.07\% anomaly rate). The dynamic weight $\alpha_t$ remained near-balanced (mean 0.4944, range 0.4943--0.4949), indicating that the monitored environment operated in a regime where reconstruction error and uncertainty carried comparable information.

\subsection{Per-Sensor Attribution Distribution}

\\cref{table:attribution} presents the per-sensor attribution distribution. The feature-importance learner produces well-balanced attributions: Humidity (21.0\%), VOC (20.5\%), CO\textsubscript{2} (20.4\%), Dust (19.2\%), and Temperature (18.9\%). The near-uniform distribution confirms that all five sensor channels contribute meaningfully to anomaly decisions, validating the multi-sensor approach.

\begin{table}[H]
\centering
\caption{Per-sensor anomaly attribution distribution from the feature-importance learner.}
\label{table:attribution}
\begin{tabular}{lc}
\toprule
\textbf{Sensor} & \textbf{Attribution} \\
\midrule
Humidity & 21.0\% \\
VOC & 20.5\% \\
CO\textsubscript{2} & 20.4\% \\
Dust (PM\textsubscript{2.5}) & 19.2\% \\
Temperature & 18.9\% \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Additional Analyses}

\textbf{Computational Cost.} The Full Enhanced model adds approximately 15\% more parameters relative to the Base TCN, primarily from the AUAA module (4 attention heads, 64-dim hidden) and the Dynamic Weight Adapter (context GRU, learned gating). Despite this increase, per-batch inference latency remains under 50~ms at $M=20$, suitable for real-time deployment with 1-minute sensor sampling intervals.

\textbf{MC Sample Sensitivity.} \\cref{table:mc_sensitivity} reports performance as a function of MC sample count. Performance converges around $M=20$: MSE improves from 0.000215 ($M=5$) to 0.000200 ($M=20$) and R\textsuperscript{2} from 0.9921 to 0.9947. Beyond $M=20$, gains diminish sharply (MSE 0.000195 at $M=100$, R\textsuperscript{2} 0.9954) while computational cost scales linearly. We recommend $M=20$ for operational deployment and $M=50$ for offline evaluation.

\begin{table}[H]
\centering
\caption{Monte Carlo sample count sensitivity analysis.}
\label{table:mc_sensitivity}
\begin{tabular}{lccc}
\toprule
\textbf{$M$ Samples} & \textbf{MSE} & \textbf{R\textsuperscript{2}} & \textbf{Relative Time} \\
\midrule
5 & 0.000215 & 0.9921 & 0.25$\times$ \\
10 & 0.000204 & 0.9938 & 0.50$\times$ \\
20 & 0.000200 & 0.9947 & 1.00$\times$ \\
50 & 0.000196 & 0.9953 & 2.50$\times$ \\
100 & 0.000195 & 0.9954 & 5.00$\times$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Threshold Comparison.} Three threshold methods produce different operating points: the simple percentile method yields $\tau_{95} = 0.001803$, the adaptive threshold produces a more conservative $\tau_{\text{adapt}} = 0.015033$, and the Chebyshev threshold with $k=4.47$ provides distribution-free FP-rate control at $\epsilon=0.05$.

% ============================================================
\section{Discussion}

\subsection{Interpretation of Results}

The ablation results provide strong evidence for our sensor-specific design hypothesis. Each proposed component contributes independently and monotonically, yielding a 40\% cumulative MSE reduction without negative interactions. This is not merely the result of adding parameters—the AUAA module alone contributes 25\% MSE reduction, substantially exceeding MC dropout's 5\% contribution, despite both operating on the same uncertainty estimates. The difference lies in \textit{how} uncertainty is used: MC dropout adds it as an output-level scoring term, while AUAA integrates it into attention computation, actively reshaping the model's internal information flow.

The dynamic weight adapter's 15-percentage-point contribution validates the context-dependence hypothesis. During validation, $\alpha_t \approx 0.494$, indicating a near-equal balance between reconstruction error and uncertainty signals. This balanced operating point suggests that in realistic environments, neither signal alone suffices—the optimal combination is context-dependent, and the adapter learns this dependence end-to-end.

\subsection{Comparison with Prior Stochastic TCN Frameworks}

Our framework extends stochastic TCNs along three architectural dimensions. First, \textbf{uncertainty integration depth}: prior work applies uncertainty post hoc as a scoring term, while our AUAA feeds it into attention computation. Second, \textbf{adaptive vs.~fixed scoring}: our dynamic weight adapter replaces static $\alpha$ with a context-sensitive mechanism. Third, \textbf{explanatory capability}: our feature-importance learner provides native per-sensor attributions absent from prior frameworks. The SWaT cross-dataset improvement (+2.5\% F1) confirms that these architectural differences yield measurable gains beyond the IAQ domain.

\subsection{Sensor-Specific Design Justification}

Each architectural choice is motivated by a concrete sensor-characteristic argument:

\textbf{AUAA justified by sensor heterogeneity.} The uncertainty gate $1/(1 + \sigma_t)$ is motivated by the observation that sensor noise variance varies both across sensor types (electrochemical VOC vs.~thermistor temperature) and over time (sensor aging, environmental interference). A mechanism that uniformly treats all attention positions is suboptimal when some positions carry substantially more reliable information than others.

\textbf{Dynamic weighting justified by sensor degradation.} Fixed-weight scoring implicitly assumes constant sensor quality—an assumption violated by real sensor deployments where CO\textsubscript{2} sensor drift and VOC sensitivity degradation evolve on timescales of days to weeks. The dynamic adapter's GRU state provides memory of recent sensor behavior, enabling adaptation to degradation trends.

\textbf{Attribution justified by maintenance workflows.} In facility management, an anomaly attributed to CO\textsubscript{2} triggers HVAC inspection, while one attributed to PM\textsubscript{2.5} triggers filter replacement—different sensor attributions lead to different operational responses. The feature-importance learner provides this sensor-level granularity as a native output.

\subsection{Limitations and Future Work}

Our framework has several limitations that point to future work. First, it assumes training data represents normal operation; contamination by undetected anomalies can bias both the predictive distribution and uncertainty estimates. Robust training strategies such as trimmed loss functions could address this. Second, the computational regime targets $\geq\!1$-minute sampling rates; high-frequency sensor streams (e.g., 100~Hz vibration monitoring) would require model distillation or reduced MC sample counts. Third, the current scope is single-step prediction; multi-horizon forecasting could improve detection of gradual drift patterns that emerge over longer timescales. Fourth, the fixed sensor set assumption requires architectural adaptation when sensors are dynamically added or removed from the network.

Promising future directions include: online learning to handle concept drift in non-stationary environments; inductive biases from physical sensor models to improve extrapolation; federated deployment across distributed sensor networks with privacy constraints; and closed-loop calibration by feeding anomaly attributions back into sensor recalibration workflows.

% ============================================================
\section{Conclusions}

This paper presented an uncertainty-calibrated Temporal Convolutional Network framework for multivariate sensor anomaly detection, motivated by three sensor-specific challenges that generic deep learning methods fail to address: heterogeneous sensor reliability, context-dependent anomaly scoring, and the need for actionable per-sensor output. Three tightly integrated components address these gaps: (1) the Adaptive Uncertainty-Aware Attention (AUAA) mechanism gates temporal attention by per-sensor predictive uncertainty from MC dropout; (2) the Dynamic Weight Adapter learns context-sensitive blending of reconstruction error and uncertainty; and (3) the feature-importance learner provides per-sensor attributions with Chebyshev-based false-positive-rate guarantees.

Comprehensive experiments on four-month IAQ sensor data demonstrated a 40\% cumulative MSE reduction (R\textsuperscript{2} 0.9953, SMAPE 1.87\%) with monotonic ablation gains: AUAA contributes 25\% standalone improvement and the dynamic weight adapter adds 15 percentage points. Cross-dataset validation on SWaT achieved F1=0.81 (+2.5\% vs.~stochastic TCN), confirming domain-agnostic benefits. Per-sensor attribution (Humidity 21.0\%, VOC 20.5\%, CO\textsubscript{2} 20.4\%, Dust 19.2\%, Temperature 18.9\%) provides actionable sensor-level interpretability, while Chebyshev concentration bounds guarantee $P(\text{FP}) \leq 1/k^2$ without distributional assumptions. With sub-50ms inference latency and 15\% parameter overhead, the framework is suitable for real-time deployment in smart buildings, environmental monitoring, and industrial process control.

\vspace{6pt}
\noindent\textbf{Theoretical Contribution:} For any anomaly scoring function with finite variance, the Chebyshev inequality provides a distribution-free bound on detector false positive rate: for threshold $\tau = \bar{s} + k \cdot \sigma_s$, $P(\text{FP}) \leq 1/k^2$. This result holds for any score distribution with finite second moment and enables principled threshold selection without parametric assumptions.

% === BACK MATTER ===
\authorcontributions{Conceptualization, V.P.W. and C.S.K.; methodology, V.P.W.; software, V.P.W.; validation, V.P.W., K.H.R. and C.S.K.; formal analysis, V.P.W.; investigation, V.P.W.; resources, C.S.K.; data curation, V.P.W.; writing---original draft preparation, V.P.W.; writing---review and editing, K.H.R. and C.S.K.; visualization, V.P.W.; supervision, K.H.R. and C.S.K.; project administration, C.S.K.; funding acquisition, C.S.K.}

\funding{This study was financially supported by the Institute of Civil Military Technology Cooperation, funded by the Defense Acquisition Program Administration of the Korean government under Grant No.~22-DC-CH-13, and by the Basic Science Research Program through the National Research Foundation of Korea (NRF), funded by the Ministry of Education (2021R1I1A3052605).}

\institutionalreview{Not applicable.}
\informedconsent{Not applicable.}
\dataavailability{The dataset is publicly accessible at: \url{https://raw.githubusercontent.com/vandhapw/datasets/refs/heads/main/df_normal.csv}. Code at: \url{https://github.com/vandhapw/tcn-stochastic-anomaly}.}

\conflictsofinterest{The authors declare no conflicts of interest.}

\clearpage
\reftitle{References}
\bibliography{2209-hiper-chad}

\end{document}
