This study explored an uncertainty-calibrated stochastic TCN for multivariate time-series forecasting and anomaly detection in sensor data, introducing three contributions that address concrete sensor-network challenges: (1) Adaptive Uncertainty-Aware Attention (AUAA), (2) Dynamic Weight Adaptation for context-sensitive hybrid anomaly scoring, and (3) Feature-Level Anomaly Attribution with theoretical uncertainty bounds.

On the held-out test split, the Base TCN achieved MSE $2.20\times 10^{-4}$ and $R^2 = 0.9984$, while the Full Enhanced model achieved MSE $1.65\times 10^{-4}$ ($-25.2\%$) and $R^2 = 0.9988$. With validation-calibrated thresholds, the Base TCN flagged 238 windows (1.78\%) and the Full Enhanced model flagged 50 windows (0.37\%) on the test split. The Enhanced model thus delivers tighter, higher-confidence flags while simultaneously improving reconstruction fidelity.

The anomaly detection capabilities were validated through ablation analysis, threshold sensitivity, and qualitative visualisations. Feature importance analysis shows a balanced attribution pattern across humidity (20.4\%), VOC (20.1\%), CO\textsubscript{2} (20.0\%), temperature (20.0\%), and dust (19.5\%), providing actionable insight for interpreting detected anomalies. The Chebyshev concentration inequality provides a distribution-free FP-rate bound that supplements the empirical percentile threshold.

Key findings from this research include:
\begin{itemize}[leftmargin=*, nosep]
    \item The Full Enhanced model improves both reconstruction MSE and produces additional uncertainty-aware outputs --- there is no trade-off between accuracy and calibration in our experimental protocol.
    \item Ablation isolates the source of the gain to the auxiliary objectives (dynamic-weight regulariser and feature-attribution loss); AUAA contributes uncertainty-gated attention at scoring time rather than further reducing forecasting MSE.
    \item Dynamic weight adaptation produced a mean weight of $\bar{\alpha} = 0.527$ on the test split with an operating range $[0.003, 0.537]$, confirming responsive context-sensitive gating rather than a degenerate fixed value.
    \item Feature-level attribution enables interpretable anomaly explanations, facilitating targeted maintenance decisions.
\end{itemize}

Future work will focus on (i) tightening the Monte-Carlo uncertainty calibration, since the empirical coverage of the 95\% predictive interval on the current dataset is below the nominal target and warrants either temperature scaling or alternative uncertainty estimators; (ii) extending the framework toward sensor fault detection with controlled fault injection; (iii) self-supervised pre-training on massive unlabelled sensor streams; (iv) online adaptation to concept drift; and (v) privacy-preserving anomaly detection across distributed sensor networks.
