\section{Discussion}
\label{sec:discussion}

The experimental results demonstrate that the proposed TCN-Stochastic framework effectively addresses the challenges of anomaly detection in multivariate indoor air quality sensor data. This section discusses the key findings, practical implications, limitations, and future research directions.

\subsection{Key Findings and Interpretation}

Our results reveal several important insights about stochastic temporal modeling for sensor anomaly detection:

\textbf{Uncertainty-Error Correlation:} The strong correlation ($r > 0.85$) between MC dropout uncertainty estimates and actual prediction errors validates the use of stochastic inference for confidence quantification in real-world sensor applications. This finding extends the theoretical work of Gal and Ghahramani \cite{gal_dropout_2016} to the practical domain of low-cost IoT sensors, where noise characteristics vary significantly across variables.

\textbf{Feature-Specific Anomaly Patterns:} The feature attribution analysis revealed that VOC and CO\textsubscript{2} are the primary drivers of detected anomalies, with distinct temporal patterns. VOC-driven anomalies typically occur during cleaning activities or chemical usage events (sharp spikes with rapid decay), while CO\textsubscript{2}-driven anomalies correlate with occupancy changes (gradual accumulation during occupied periods). This interpretability addresses a critical limitation of black-box deep learning models in safety-critical monitoring applications.

\textbf{Architectural Efficiency:} The ablation study confirms that each component of our architecture contributes meaningfully to performance. The 44.9\% MSE degradation when removing dilated convolutions underscores their importance for capturing multi-scale temporal dependencies in sensor data. Similarly, the 18.5\% performance drop without MC dropout demonstrates the value of uncertainty-aware predictions for distinguishing true anomalies from sensor noise.

\subsection{Practical Implications for IAQ Monitoring}

The computational efficiency analysis positions TCN-Stochastic as a viable solution for real-time deployment:

\textbf{Edge Deployment Feasibility:} With a model size of only 210 KB and inference latency of 0.8 ms per batch, the framework can be deployed on resource-constrained edge devices such as Raspberry Pi or ESP32 microcontrollers with external memory. This enables localized anomaly detection without requiring continuous cloud connectivity, addressing privacy concerns and reducing bandwidth costs for large-scale sensor networks.

\textbf{Proactive Maintenance:} The 5.0\% anomaly detection rate, combined with feature attribution, enables facility managers to identify specific intervention needs. For example, VOC-driven anomalies may trigger ventilation system activation, while CO\textsubscript{2}-driven anomalies could prompt occupancy management or HVAC optimization. The uncertainty estimates provide confidence scores that help prioritize responses, reducing false alarm fatigue.

\textbf{Sensor Health Monitoring:} Beyond environmental anomaly detection, the framework shows promise for sensor fault diagnosis. Persistent high uncertainty in specific variables may indicate sensor drift or calibration issues, enabling predictive maintenance before complete sensor failure.

\subsection{Comparison with Prior Work}

Our results align with and extend recent studies on stochastic TCNs:

\textbf{vs. Electricity Demand Forecasting:} Ghimire et al. \cite{ghimire_uncertainty_2024} applied TCN with MC dropout for electricity demand uncertainty modeling, achieving similar loss reduction patterns. However, their single-variable setting differs fundamentally from our multivariate approach, where inter-sensor correlations introduce additional complexity. Our hybrid scoring mechanism specifically addresses this by weighting reconstruction errors across correlated variables.

\textbf{vs. Hydrological Forecasting:} Le et al. \cite{le_streamflow_2025} demonstrated MC dropout for streamflow forecasting, reporting comparable uncertainty calibration. The key distinction lies in temporal dynamics: hydrological data exhibits slow, seasonal variations (daily to yearly scales), whereas IAQ sensor data combines rapid fluctuations (seconds to minutes for VOC spikes) with gradual trends (temperature drift over hours). Our exponentially increasing dilation rates ($d=1,2,4$) are specifically optimized for this multi-scale characteristic.

\textbf{vs. Calibrated SDE Approaches:} Cui et al.'s \cite{cui_sde-hnn_2025} SDE-HNN framework provides theoretically grounded uncertainty calibration through stochastic differential equations. While their approach offers stronger mathematical guarantees, our MC dropout method achieves competitive performance with significantly lower implementation complexity and computational overhead, making it more accessible for practical IoT deployments.

\subsection{Limitations}

Despite promising results, several limitations warrant acknowledgment:

\textbf{Single-Dataset Validation:} The model was evaluated exclusively on indoor air quality data from one laboratory environment over a 4-month period. While comprehensive (178,560 readings, 5 variables), this limits claims about generalizability to other domains such as industrial monitoring, healthcare sensors, or smart city infrastructure. Cross-domain validation remains an important direction for future work.

\textbf{Unsupervised Setting:} Our anomaly detection operates in an unsupervised manner, learning normal patterns without labeled anomalies. While this enhances practical applicability (labels are often unavailable), it prevents direct computation of precision, recall, and F1 scores. The 5.0\% detection rate is based on percentile thresholds rather than ground truth validation. Future work should incorporate expert-labeled anomalies for supervised evaluation.

\textbf{Static Threshold Selection:} Although we employ adaptive multi-percentile thresholds ($\tau_{95}$ and $\tau_{99}$), these are determined offline from historical error distributions. Dynamic threshold adaptation to non-stationary environmental changes (e.g., seasonal shifts, building renovations) remains an open challenge. Online learning approaches could address this limitation.

\textbf{Computational Overhead of MC Dropout:} While the absolute inference time (0.8 ms) meets real-time requirements, the MC dropout mechanism requires 50 forward passes per prediction, introducing a $50\times$ overhead compared to deterministic inference. For ultra-low-latency applications (sub-millisecond requirements), alternative uncertainty estimation methods such as deep ensembles or evidential deep learning may offer better efficiency trade-offs.

\subsection{Future Research Directions}

Based on the findings and limitations discussed, we propose several avenues for future investigation:

\textbf{Multi-Domain Validation:} Extending evaluation to diverse sensor domains (industrial IoT, healthcare wearables, autonomous vehicles) would establish the generalizability of the TCN-Stochastic framework. Each domain presents unique challenges: industrial sensors exhibit higher noise levels, healthcare data requires stricter privacy guarantees, and automotive applications demand ultra-low latency.

\textbf{Self-Supervised Learning Integration:} Incorporating self-supervised pre-training on unlabeled sensor data could improve representation learning, especially in scenarios with limited labeled examples. Contrastive learning approaches tailored to temporal sequences (e.g., TS-TCC, CoST) may enhance the model's ability to distinguish normal variations from true anomalies.

\textbf{Federated Learning for Privacy-Preserving Deployment:} For applications spanning multiple buildings or organizations, federated learning would enable collaborative model training without sharing raw sensor data. The compact model size (210 KB) makes TCN-Stochastic well-suited for federated settings with bandwidth constraints.

\textbf{Causal Anomaly Attribution:} While current feature attribution identifies which variables contribute to anomalies, causal inference methods could determine why anomalies occur by modeling causal relationships between sensors. Integration with structural causal models (SCMs) or Granger causality analysis would enable root-cause diagnosis beyond correlation-based attribution.

\textbf{Continuous Learning and Adaptation:} Developing online learning mechanisms would allow the model to adapt to gradual environmental changes (sensor drift, seasonal patterns, building modifications) without catastrophic forgetting. Elastic weight consolidation (EWC) or memory-augmented approaches could preserve learned normal patterns while accommodating legitimate distribution shifts.

\subsection{Summary}

In summary, the TCN-Stochastic framework demonstrates strong performance for multivariate time-series anomaly detection in indoor air quality monitoring, with practical advantages in computational efficiency and interpretability. While limitations in dataset diversity and unsupervised evaluation warrant further investigation, the results establish a solid foundation for real-world deployment and future research extensions.
