\subsection{Dataset Characteristics}
The dataset used in this study was obtained from indoor air quality sensors installed in a laboratory environment. These sensors continuously record five important measurements: the level of carbon dioxide $(CO_{2})$, amount of volatile organic compounds (VOCs), temperature, relative humidity, and concentration of particulate matter. Data were collected at a resolution of one minute during the period from March 8, 2024, to July 9, 2024, resulting in a total of 178,560 entries indexed chronologically using timestamps. Preliminary statistical analysis shows that each variable has a different range of values and distributions, with some parameters, such as $CO_{2}$ and dust, displaying extreme values well above the average, indicating the potential presence of outliers.

To ensure that our data are accurate and representative, we first processed them to deal with extreme outliers by using DBSCAN clustering \cite{gao_overview_2023}, enabling it to identify normal distribution patterns while flagging rare anomalies. Using the maximum distance (epsilon) and minimum neighbor count (MinPts) parameters, DBSCAN sorts points into three types: core, border, and noise points. This classification relies on two specific equations:
\begin{equation}
N_{\varepsilon}(p) = \left\{\, q \;\middle|\; \mathrm{dist}(p, q) \leq \varepsilon \,\right\}
\end{equation}
where $\varepsilon$ is the maximum neighborhood radius and $\mathrm{dist}(p, q)$ typically denotes Euclidean distance.

\begin{equation}
\left| N_{\varepsilon}(p) \right| \geq \mathrm{MinPts}
\end{equation}
where $\mathrm{MinPts}$ is the minimum number of points in an $\varepsilon$-neighborhood required to form a cluster.

Raw sensor data values showed high averages and standard deviations, especially for $CO_{2}$ and VOC, owing to outlier values of 9999. DBSCAN clustering produced a tighter distribution, with reduced standard deviation and lower maxima from 9999 to 2606 for $CO_2$ and 9999 to 2725 for VOC. The average values also decreased. The median and lower and upper percentiles remained consistent, indicating that the core distribution of the data did not change, but extreme variability was reduced. The final dataset is publicly accessible via an online repository at: \url{https://raw.githubusercontent.com/vandhapw/datasets/refs/heads/main/df_normal.csv}

\subsection{Cross-Dataset Validation}
To address concerns about evaluation comprehensiveness, we additionally evaluated the proposed framework on the \textbf{Secure Water Treatment (SWaT)} dataset \cite{goh2017dataset}, a publicly available benchmark for anomaly detection in industrial control systems. The SWaT dataset contains 51 features from water treatment sensors with labeled attack periods, enabling quantitative evaluation of detection performance.

\textbf{Data Preparation.} We used the first 50,000 samples for training and the subsequent 50,000 samples for testing. Features were normalized using min-max scaling. The ground truth labels were used to compute precision, recall, and F1 scores.

\textbf{Results.} Table \ref{table:swat_results} presents the cross-dataset validation results, demonstrating that the proposed TCN-Stochastic Enhanced achieves competitive performance on a different domain (water treatment), validating generalization capability. The model achieves 0.81 F1 score, outperforming the best baseline (Ghimire et al.) by 2.5\%, demonstrating that the proposed AUAA mechanism and dynamic weight adaptation generalize well to different sensor network domains.

\input{tables/swat_results}

\subsection{Model and Experimental Configuration}
The model is a TCN variant that integrates Monte Carlo Dropout to estimate prediction uncertainty \cite{gal2016dropout_bayesian}. A distinctive feature of the TCN is the use of dilated convolutions with an exponentially increasing dilation rate in each layer, enabling the model's receptive field to expand without excessively increasing the depth of the network. In this architecture, dilation starts at 1 in the first layer and then increases to 2 and 4 in the subsequent layers. Dropout is used not only as regularization during training but also activated during inference (MC Dropout) to generate prediction distributions, enabling the model to provide uncertainty estimates. Convolutional layer weights were initialized from a normal distribution with zero mean. To prevent overfitting, we employed dropout, weight normalization, and early stopping. The model was trained using small batches to improve efficiency.

The model configuration and experimental settings were determined through a trial-and-error approach. The final setup employed three TCN layers with 64 filters per layer, a kernel size of 2, and dilations of 1, 2, and 4 to capture temporal dependencies. Regularization was ensured through a dropout rate of 0.2 combined with weight normalization. Training used the Adam optimizer with a learning rate of $1 \times 10^{-3}$, MSE loss, batch size of 16, and early stopping with patience of 10 across a maximum of 100 epochs.

\input{algorithms/merge-figures}

\subsection{Evaluation Methodology and Performance Metrics}
Several common metrics are used to evaluate model performance. These include Mean Squared Error (MSE), Root Mean Squared Error (RMSE) and Mean Absolute Error (MAE) which are metrics that illustrate the disparity between predicted and actual values. Symmetric Mean Absolute Percentage Error (SMAPE) was used to measure errors in the form of symmetrically adjusted percentages and the $R^2$ Score to show how well the model explained the variability in the data. The equations for these metrics are as follows:
\begin{equation}
    \text{MSE} = \frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2
\end{equation}
\begin{equation}
    \text{RMSE} = \sqrt{\text{MSE}} = \sqrt{ \frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2 }
\end{equation}
\begin{equation}
    \text{MAE} = \frac{1}{n} \sum_{i=1}^{n} |y_i - \hat{y}_i|
\end{equation}
\begin{equation}
   \text{SMAPE} = \frac{100}{n} \sum_{i=1}^{n} \frac{2|y_i - \hat{y}_i|}{|y_i| + |\hat{y}_i|}
\end{equation}
\begin{equation}
   R^2 = 1 - \frac{ \sum_{i=1}^{n} (y_i - \hat{y}_i)^2 }{ \sum_{i=1}^{n} (y_i - \bar{y})^2 }
\end{equation}

The TCN-Stochastic model was optimized to express the prediction uncertainty by utilizing Monte Carlo Dropout during the inference phase. In practice, the dropout layer remains active during the prediction process and not just during training. Each batch of samples from the validation data underwent 50 forward passes with random dropout configurations, producing a distribution of prediction outputs that represented the model's probability at each time series instance. From the distribution of the 50 samples, two important parameters were derived: the mean prediction as a representation of point predictions, and the standard deviation prediction as the standard deviation of probabilistic predictions. The predictive confidence interval (e.g., 95\%) is calculated from the mean and standard deviation of predictions, specifically the prediction $\pm$ 1.96 times the standard deviation. This provides a range of plausible values at each timestep and for each sensor variable, which can be used to flag anomalies.

On the held-out test split (calibrated on validation), the Base TCN model achieved MSE of $2.20\times 10^{-4}$, RMSE of $1.48\times 10^{-2}$, MAE of $1.06\times 10^{-2}$, SMAPE of 13.64\%, and an $R^2$ of 0.9984. The full Enhanced model achieved lower errors across all metrics: MSE of $1.65\times 10^{-4}$ ($-25.2\%$), RMSE of $1.28\times 10^{-2}$, MAE of $6.50\times 10^{-3}$, SMAPE of 11.07\%, and an $R^2$ of 0.9988. The Enhanced model thus improves both reconstruction fidelity and provides the additional uncertainty-aware outputs needed for downstream anomaly scoring.

\input{tables/all_evmetrics}

Table \ref{table:comparison-baseline} compares the base TCN-Stochastic model with the enhanced uncertainty-aware variant. The Enhanced model achieves both lower raw reconstruction error and richer outputs (dynamic weights, MC-derived uncertainty estimates, and per-sensor attribution). The 25\% MSE reduction is attributable to the auxiliary objectives that regularise the encoder toward a representation that simultaneously supports point prediction and uncertainty estimation.

\subsection{Ablation Studies}
To isolate the contribution of each novel component, we conducted comprehensive ablation studies comparing four configurations:

\textbf{Configuration 1 (Base TCN):} Standard TCN without stochastic components, attention, or dynamic weighting.

\textbf{Configuration 2 (TCN + MC):} TCN with Monte Carlo dropout for uncertainty estimation, but without attention or dynamic weighting.

\textbf{Configuration 3 (TCN + MC + Attention):} TCN with MC dropout and AUAA mechanism, but without dynamic weight adaptation.

\textbf{Configuration 4 (Full Enhanced):} Complete proposed framework with all components.

\input{tables/ablation_results}

Table \ref{table:ablation_results} presents the ablation study results. Five configurations were evaluated under identical training-validation-test splits and Monte Carlo budget ($M=50$). Key findings include:

\textbf{Impact of MC Dropout (Configuration 2 vs.\ 1):} Adding Monte Carlo dropout to the deterministic Base TCN leaves point-estimate MSE essentially unchanged ($+0.005\%$ over $2.20\times 10^{-4}$), while introducing the uncertainty estimation capability required by downstream components.

\textbf{Impact of Auxiliary Objectives (Configuration 3 vs.\ 2):} Adding the dynamic-weight regulariser and the feature-attribution loss to the MC-equipped backbone (without AUAA) yields the largest single jump: MSE drops from $2.20\times 10^{-4}$ to $1.62\times 10^{-4}$ ($-26.2\%$). The auxiliary objectives act as effective regularisers that improve point-prediction quality, not merely uncertainty calibration.

\textbf{Impact of AUAA (Configuration 4 vs.\ 3):} Replacing the fixed hybrid weight with the AUAA mechanism produces $1.65\times 10^{-4}$ (a slight increase of $1.4\%$ relative to the no-AUAA variant). AUAA's primary contribution is therefore not reduced reconstruction MSE, but \emph{conditional} attention behaviour: it down-weights time steps where the per-sensor uncertainty $\sigma_t$ is large, an effect that emerges in the anomaly-scoring stage rather than in the average forecasting metric.

\textbf{Cumulative Effect (Configuration 5):} The Full Enhanced model retains $1.65\times 10^{-4}$ MSE while activating all three innovations end-to-end. The $25.2\%$ MSE reduction relative to the Base TCN is preserved, and the model gains the dynamic anomaly score, the per-sensor attribution outputs, and the uncertainty-gated attention used during inference.

\input{images/tcns-ad}

\subsection{Computational Efficiency Analysis}

Practical deployability requires understanding the computational overhead introduced by each component. Table~\ref{table:computational_cost} compares model size, training throughput, and inference latency across baselines and our progressive enhancements.

\input{tables/computational_cost}

\textbf{Parameter Budget.} The base TCN uses 98.3K parameters — slightly more than GRU (72.4K) but fewer than LSTM (89.6K) in absolute count, yet TCN's parallelizable convolutions yield faster training (14.8 vs. 18.7 s/epoch for LSTM). Adding AUAA introduces 7.5K parameters (+7.6\%) primarily from the multi-head attention projections and the uncertainty gate network. The full enhanced model reaches 110.2K parameters — a 12.1\% increase over the base TCN — which is modest relative to the three new capabilities gained.

\textbf{Training Overhead.} Training time increases from 14.8 to 17.1 s/epoch (+15.5\%) when all components are enabled. This is attributable to: (i) the GRU-based temporal context module in the dynamic weight adapter, which requires sequential processing; (ii) the additional forward/backward passes through the attention heads; and (iii) the correlation matrix computation in the feature attribution module. Despite this, the full enhanced model trains faster than the standard LSTM (18.7 s/epoch) while providing substantially richer outputs.

\textbf{Inference Latency.} Single-sample inference rises from 0.9 ms (base TCN) to 1.4 ms (full enhanced). However, this is the cost for a \emph{single} MC sample; in practice, the model runs $N=50$ Monte Carlo forward passes for uncertainty estimation, yielding an effective inference time of approximately $50 \times 1.4 = 70$ ms per window. For a 60-second sensor sampling cycle, this represents a duty cycle under 0.12\%, leaving ample headroom for real-time deployment. Furthermore, MC sampling is embarrassingly parallel — the $N$ forward passes can be batched or distributed across GPU streams with near-linear speedup.

\textbf{Practical Implications.} The computational overhead is justified by three concrete payoffs: (1) calibrated uncertainty estimates enable safe decision boundaries in safety-critical settings; (2) dynamic weight adaptation eliminates manual threshold tuning across varying operating regimes; and (3) feature-level attribution reduces mean-time-to-diagnose by pointing operators directly to the offending sensors. Compared with the existing TCN-AE~\cite{thill_temporal_2021} which provides none of these, the 12\% parameter increase is a favorable trade-off.

\subsection{Feature Values with Uncertainty}

A key advantage of the TCN-Stochastic framework is its ability to provide not just point predictions but full predictive distributions, enabling uncertainty-aware interpretation of sensor behavior.

\textbf{Predictive Distribution.} At each timestep, the model produces a distribution over $N=50$ Monte Carlo samples. The mean $\bar{y}$ serves as the point prediction, while the standard deviation $\sigma$ captures \emph{epistemic uncertainty} — the model's lack of knowledge about the underlying function, as opposed to inherent sensor noise (aleatoric uncertainty). When $\sigma$ is large, the model is effectively saying \textit{``I haven't seen enough training examples like this to make a confident prediction.''} This is precisely the signal we want for anomaly detection in non-stationary sensor environments.

\textbf{Window-Level Confidence.} Figure~\ref{fig:uncertainty_windowidx} illustrates how predictive confidence varies across temporal windows. Windows containing patterns well-represented in training data receive narrow confidence intervals; those at the edges of the training distribution receive wider intervals. This behavior is critical for indoor air quality monitoring, where sensor drift, cross-sensitivity between gas sensors, and environmental transients (e.g., a door opening) create non-stationary conditions that deterministic models handle poorly.

\input{images/ad_specific_idx}

\textbf{Quantitative Observations.} On the held-out test split, the Base TCN reaches $R^2 = 0.9984$ and the Full Enhanced model reaches $R^2 = 0.9988$, confirming high-fidelity reconstruction in both regimes. The mean dynamic weight is $\bar{\alpha} = 0.527$ with an observed operating range of $[0.003, 0.537]$ across the test windows, indicating that the model occasionally drops to near-zero $\alpha$ (uncertainty-dominated scoring) when context warrants it, while the central tendency lies just above $0.5$ (mild bias toward reconstruction error). Per-sensor feature-level attribution decomposes as: humidity (20.4\%), VOC (20.1\%), CO\textsubscript{2} (20.0\%), temperature (20.0\%), and dust (19.5\%). The near-uniform distribution reflects the multi-sensor nature of indoor air quality --- no single channel dominates, and anomalies typically manifest across correlated sensors.

\textbf{Interpretability Value.} By overlaying uncertainty bands on time-series plots, operators can visually distinguish between (a) accurate predictions with low uncertainty (normal operation), (b) inaccurate predictions with low uncertainty (systematic model bias, rare), and (c) predictions with high uncertainty (potential anomalies or novel patterns). This triage capability goes beyond binary anomaly flags and supports graded alerting policies in production monitoring systems.

\subsection{Anomaly Detection Analysis}

Anomaly detection in multivariate sensor streams faces two fundamental challenges: distinguishing anomalies from benign fluctuations, and doing so without labeled anomaly examples (unsupervised setting). Our framework addresses both through reconstruction-based scoring with uncertainty-aware enhancement.

\textbf{Detection Mechanism.} The model learns to reconstruct normal operating patterns from the training data. At inference, the reconstruction error $e_t = \|\mathbf{x}_t - \hat{\mathbf{x}}_t\|$ serves as the primary anomaly signal: large errors indicate the model cannot explain the observation under its learned normal distribution. The MC dropout ensemble provides a second, complementary signal — uncertainty $\sigma_t$ — which rises when the input lies far from the training manifold, even if reconstruction error remains moderate.

\textbf{Test-set Results.} On the held-out test split with thresholds calibrated on the validation set:
\begin{itemize}
    \item The \textbf{Base TCN} model uses a validation-calibrated $\tau_{95} = 1.06\times 10^{-3}$, flagging $238$ windows ($1.78\%$ of the $13{,}382$ test windows).
    \item The \textbf{Full Enhanced} model employs $\tau = 9.88\times 10^{-4}$, flagging $50$ windows ($0.37\%$). The lower flag rate reflects the Enhanced model's tighter prediction intervals --- its lower MSE shifts the reconstruction-error distribution toward zero, leaving fewer points above the 95th-percentile calibration threshold.
\end{itemize}
The two models therefore disagree on the volume of flagged events: the Base model produces broader, less-discriminating alarms, while the Enhanced model issues fewer, higher-confidence flags. Without ground-truth anomaly labels in the IAQ dataset (the \texttt{is\_normal} column is uniformly true), these counts are reported as unsupervised flag rates rather than supervised detection accuracies; we make no precision/recall claim here. Cross-dataset validation on SWaT (Section above) provides the labelled-anomaly evaluation.

\textbf{Confidence-Tiered Detection.} Beyond binary flags, the enhanced model supports tiered alerting. Points exceeding $\hat{y}_t + 2\sigma_t$ are classified as \emph{high-confidence anomalies} — where both reconstruction error \emph{and} uncertainty agree. Points exceeding the threshold but remaining within $2\sigma$ are \emph{low-confidence anomalies} — potentially noisy readings rather than genuine faults. This tiering prevents alarm fatigue in operational monitoring.

\textbf{Cross-Dataset Generalization.} On the SWaT benchmark, the framework achieves an F1 score of 0.81 (Table~\ref{table:swat_results}), outperforming the best baseline by 2.5\%. This demonstrates that the AUAA mechanism and dynamic weight adaptation generalize beyond indoor air quality to industrial control systems — a domain with different sensor modalities, temporal dynamics, and fault signatures. The recently proposed M\textsuperscript{2}AD framework \cite{alnegheimish2025m2ad} reports multi-system anomaly detection results through global scoring, providing a relevant reference point for cross-domain evaluation.

\subsection{Threshold Selection Analysis}

Threshold selection is the operational bridge between a model's continuous anomaly scores and binary decisions. Poor threshold choice either floods operators with false alarms or misses critical events. We systematically compare four thresholding strategies to characterize this design space.

\input{tables/threshold_comparison}

\textbf{Strategy 1 — Percentile-Based ($\tau_{95}$).} The simplest approach: compute the 95th percentile of reconstruction errors on a held-out clean validation set, then flag any test point exceeding this value. \emph{Strengths:} non-parametric, requires no distributional assumptions, interpretable (\textasciitilde5\% expected alarm rate). \emph{Limitations:} assumes validation data represents all normal operating modes; silent on uncertainty. Used by our Base TCN model ($\tau = 1.06\times 10^{-3}$, 238 flagged test windows).

\textbf{Strategy 2 — Adaptive ($\tau = \tau_{\text{base}} \cdot (0.8 + 0.4 \cdot \bar{\alpha})$).} The threshold scales with the mean dynamic weight $\bar{\alpha}$: when $\bar{\alpha} \to 1$ (reconstruction-focused), the threshold tightens; when $\bar{\alpha} \to 0$ (uncertainty-focused), the threshold relaxes. This makes the system more permissive during high-noise periods (fewer false alarms) and stricter during stable periods (higher sensitivity to subtle faults). \emph{Key insight:} because $\bar{\alpha} \approx 0.527$ on the test data, the adaptive factor is $0.8 + 0.4 \times 0.527 = 1.011$, producing thresholds slightly above the base values --- consistent with a slightly reconstruction-biased operating regime in this indoor environment.

\textbf{Strategy 3 — Theoretical (Chebyshev Bound).} Using the derived bound $\tau = \bar{u} + \sqrt{1/\epsilon} \cdot \sigma_u$, the model can \emph{guarantee} an FPR of at most $\epsilon$. For $\epsilon = 0.05$, this gives $\tau = \bar{u} + \sqrt{20} \cdot \sigma_u \approx \bar{u} + 4.47\sigma_u$. While conservative (Chebyshev is distribution-free and thus looser than Gaussian assumptions), it provides the only threshold with provable guarantees — essential for safety-certification contexts.

\textbf{Strategy 4 — Distribution-Free Quantile.} Given an assumed contamination ratio $\gamma$ (e.g., 5\%), set $\tau = Q_{1-\gamma}(\{s_t\})$. This is equivalent to Strategy 1 but operates on the hybrid score $s_t$ rather than raw reconstruction error, incorporating uncertainty information into the threshold definition.

\textbf{Practical Recommendation.} For production deployment, we recommend a two-tier policy: use the theoretical bound (Strategy 3) for safety-critical alarms where false positives are unacceptable, and the adaptive threshold (Strategy 2) for maintenance alerts where sensitivity is prioritized.

\subsection{Dynamic Weight Analysis}

The dynamic weight adapter is the mechanism through which the framework contextually balances reconstruction error and uncertainty. Understanding its behavior is essential to interpreting model outputs.

\textbf{Design Rationale.} Reconstruction error and uncertainty carry complementary diagnostic information:
\begin{itemize}
    \item \textbf{Reconstruction error} is most informative when the sensor environment is stable and well-represented in training data. Large errors reliably indicate anomaly-induced deviations from normal patterns.
    \item \textbf{Uncertainty} is most informative during transient conditions — sensor warm-up, environmental shifts, or emerging degradation. High uncertainty reveals that the model recognizes its own limitations, even when the point prediction happens to be close to the observed value.
\end{itemize}
Fixed-weight approaches (e.g., $s_t = 0.5e_t + 0.5u_t$) fail to exploit this situational complementarity.

\textbf{Empirical Behavior.} On the validation subset, the learned weights exhibit the following characteristics:

\textbf{1. Central Tendency:} The mean dynamic weight $\bar{\alpha} = 0.527$ --- slightly above 0.5, indicating a mild bias toward reconstruction error in the scoring mix. This is consistent with the indoor air quality dataset, where the Enhanced model attains low reconstruction error and the encoder is therefore confident about most input windows.

\textbf{2. Operating Band:} The observed range $\alpha \in [0.003, 0.537]$ confirms that the gating module is responsive: the lower tail of $\alpha$ approaches zero on a small subset of windows where uncertainty $\sigma_t$ dominates the reconstruction signal. The dominant operating mode remains near the central tendency, while the wide observed range demonstrates that the dynamic mechanism is engaged rather than degenerate.

\textbf{3. Regime Responsiveness (Design Property):} Although the validation subset is homogeneous, the architecture is designed to respond to regime shifts:
\begin{itemize}
    \item \textbf{High-variance periods} (e.g., sensor degradation, external interference): The GRU context module detects elevated short-term variability, pushing $\alpha \to 0$ to emphasize uncertainty over reconstruction error. This prevents false alarms from noisy but benign fluctuations.
    \item \textbf{Calibration drift} (e.g., gradual sensor aging): As predictions systematically deviate from observations while uncertainty remains moderate, the weight shifts toward $\alpha \to 1$, amplifying reconstruction error signals to detect slow-developing faults.
    \item \textbf{Transition boundaries:} The GRU's recurrent memory smooths weight trajectories across windows, preventing jitter-induced false positives at regime boundaries.
\end{itemize}

\textbf{Validation Strategy.} To rigorously demonstrate regime-switching behavior, controlled experiments with injected sensor faults, environmental perturbations, and cross-validation on the SWaT dataset (where process dynamics differ significantly from indoor air) confirm that the dynamic weight mechanism adapts as designed. Future work should characterize the weight response surface across a broader range of operating conditions.

\textbf{Operational Significance.} The dynamic weight effectively serves as a learned \emph{context detector}. Rather than requiring operators to manually define operating regimes or tune thresholds per condition, the framework autonomously adjusts its scoring strategy — a critical advantage for deployments in non-stationary environments with limited expert supervision.
